\textbf{Real-time guitar transcription using\\Fourier transform based methods}\\A pitch estimation framework and overtone sieve algorithm

### 2021-04-14 paper.tex
fixed small mistakes

\section*{Abstract}
In this paper, we explore real-time transcription of a electric guitar signal. Data driven methods are not fast enough for real-time use, so we focus on Fourier transform based methods. We will provide a basis for monophonic transcription and a proof-of-concept algorithm for polyphonic transcription. The goal of this paper is to give an overview of the problems when trying to solve this problem.
% In the field of automatic music transcription (AMT), data driven methods such as artificial intelligence and machine learning are becoming more popular by the year. This is especially true for multi-pitch estimation (also known as multiple $f_0$ estimation). Even though the results are getting better every year, little understanding of the problem is gained. The few papers which use mathematical models to solve this problem, do not publish their code which makes it difficult to build upon.

@@

\section{Introduction}
Automatic Music Transcription (AMT) is a field of study which tries to convert acoustic recordings into some form of (digital) music notation~\cite{survey1}. This task can be divided into two subproblems, pitch estimation and onset detection~\cite{survey2}. Instead of trying to convert a song into music notation, this paper will focus on translating a guitar signal into a set of played notes in real-time. This problem is very similar to AMT, as both require pitch estimation and onset/offset detection, however, in our case we also have take latency into account.

Single pitch estimation can be accurately performed using simple mathematical models~\cite{mono}. Multi pitch estimation is much harder as harmonic overtones are difficult to differentiate from the fundamental notes~\cite{oud}. Because of this, most recent methods use machine learning or artificial intelligence. These methods are not well suited for real-time transcription. Also, no actual (mathematical) understanding of the problem is gained. Because of this, this paper will focus on mathematically solving this problem. Even if the methods researched in this paper might not outperform ML/AI solutions by themselves, they could be used to filter the input data for the ML/AI approaches. This might in turn improve the results of the AI/ML methods significantly.

@@

A big problem in polyphonic pitch estimation is the occurrence of overtones~\cite{oud}. As these overtones also have a periodic nature in the frequency domain, they can be detected using another Fourier transform~\cite{double}. This does however have a big impact on the latency, as performing a Fourier transform is computationally expensive. Also, notes which are at octave intervals have the same periodicity~\cite{octave}, which makes them difficult to distinguish. This is called the octave problem.

Another problem is choosing the real-time constraint for processing the input. Some papers claim 140 ms is sufficient~\cite{sloomboi}. This is however ridiculously long, as it would be about a note difference when playing eight notes in 200 BPM. Also, as there is already a large latency due to how audio drivers work~\cite{os}. In practice, musicians often tweak their audio driver settings in order to gain 5 to 10 ms in latency, which makes the difference between being able to play faster pieces or not.

@@

As stated before, our goal is to perform the transcription in real-time. We choose a constraint of 6 ms. Most musicians work with a driver latency of 9 to 11 ms. On a good computer, this latency can be reduced to 3-5 ms. We will use these 6 ms of latency difference to perform our calculations in. Note that when playing more notes per second, latency becomes more of a problem. On slower pieces, we can get away with much larger latencies. In order to test this for yourself, we developed a tool which plays back the incoming audio signal with a user specifiable delay. This tool can be found under the "latency" branch in our GitHub repository.

@@

\subsection{Transform to frequency domain}
To estimate the frequency components of a frame, transforms are used. Most commonly, the Fourier transform is used as it is the most researched transform and much is known about it~\cite{survey2}. Its main downside is that the frequency bins are constant is size, where the distance between notes increases exponentially. Because of this, low notes are hard to discern in the frequency domain where high notes have more resolution than needed. Because of this, other transforms such as the constant Q transform (CQT) are getting more popular.

When using these transforms, window functions have to by applied to the frames to prevent spectral leakage. Furthermore, the actual peak locations can be accurately interpolated using (Lagrangian) quadratic interpolation~\cite{interpol}. These concepts are further described in Section~\ref{sub:four} and Section~\ref{sub:peak} respectively.

\subsection{Peak picking}
Peaks in the frequency domain correspond to predominant frequencies in the analysed frame. However, not every peak is of interest. Some peaks are generated by noise. Many of these peaks can be filtered by requiring a minimal signal level to be considered a peak. Furthermore, due to spectral leakage, peaks can be observed which are not in the original signal. This kind of noise can be minimized by choose a good window function~\cite{window}. In polyphonic signals, harmonic overtones may slightly interfere, which will may lead to inconsistent and oscillating peaks.  % TODO: Maybe make preliminary section about overtone dissonance

@@

\section{Implementation}  \label{sec:impl}
The program developed for this paper is written in C++ (11) and compiled with g++ (10.2.0). It uses SDL (2.0.14) for audio input/output and spectrum visualisation. FFTW (3.3.9) is used for efficient Fourier transforms.

There are many parameters which change the behaviour of the overall system. For this paper, we choose some parameters which seemed perform well, however, these parameters can likely be optimized. All parameters can be found in \texttt{config.h} along with comments explaining them.

@@

Before applying a transform, a window function is often applied to minimize the artefacts produced by the transform. In this paper, we mainly use the Blackman-Nuttall window, but the following windows are also provided: Hamming window, Hann window, Blackman window, Nuttall window, Blackman-Harris window, Flat top window. Other window functions can easily be added.

\subsection{Peak picking} \label{sub:peak}
Three different peak picking algorithms have been implemented. The first picks every local maximum (every \texttt{i} where \texttt{data[i - 1] < data[i] \&\& data[i] > data[i + 1]}). The other two algorithms pick peaks based on a Gaussian average envelope. One algorithm picks all peaks which are above this average and the other only picks the highest peak for each spectral lobe above this average. Using this envelope, we effectively filter small local maxima close to large local maxima.

The Gaussian average envelope is calculated for every Fourier bin $i$ by taking the weighted average of a arbitrary number of bins around $i$. The weights are determined by a Gaussian function ($e^{-\pi(\frac{x}{\sigma})^2}$). The value of sigma and number of bins around $i$ can be changed in \texttt{config.h}. Higher values for $\sigma$ makes points around the considered point have a higher weighting. We choose a value of 1.2, but this value could be optimized for better results.

Even though this envelope works well for parts of the spectrum with peaks, it will still find peaks in noisy areas without peaks. This can be prevented in two ways. The first is setting a threshold which a peak has to exceed. This will result in less sustain, but works well. Secondly, we know that peaks outside the frequency range of a guitar cannot be a fundamental frequency. We can discard peaks lower than the lowest note, but peaks higher than the highest note still provide us information as will be described in Section~\ref{sub:impnoteset}

@@

We use a frame size of 16384 samples and a sample rate of 192000 samples per second, which equates to a frame length of 85.3 ms. This is well over our real-time constraint. Unfortunately, we cannot change this value too much. The sample rate is already at the maximum of most audio interfaces. We also cannot lower the number of samples per frame while using the Fourier transform, as it would further reduce the frequency resolution. However, as will be explained in Section~\ref{sub:futlaten}, this might not be a problem, as we can reduce this latency by employing multi-threading.

@@

We also constructed our own dataset to evaluate the performance. This dataset will be a more technical dataset; consisting of scales and intervals played on a guitar. Scales allow us to measure the monophonic performance. As scales are easy to play, multiple recordings should have the same results. This allows us to isolate and measure other variables, such as the effect of tempo or playing the scale higher/lower. Recordings of different intervals allow us to measure the polyphonic performance. By measuring all two note intervals, we can identify any problems with specific intervals. As the distance in frequency in an interval increases when the same interval is played at higher notes, we record the intervals at multiple root notes. We will also measure the performance on the four basic triads: the major triad, minor triad, augmented triad and diminished triad.

>>>>

\section*{Abstract}
In this paper, we explore real-time transcription of a electric guitar signal. Data driven methods are not fast enough for real-time use, so we focus on Fourier transform based methods. We will provide a basis for monophonic transcription and a proof-of-concept algorithm for polyphonic transcription. The goal of this paper is to give an overview of the problems when trying to solve this problem and to provide an implementation in which different methods can be tested.
% In the field of automatic music transcription (AMT), data driven methods such as artificial intelligence and machine learning are becoming more popular by the year. This is especially true for multi-pitch estimation (also known as multiple $f_0$ estimation). Even though the results are getting better every year, little understanding of the problem is gained. The few papers which use mathematical models to solve this problem, do not publish their code which makes it difficult to build upon.

@@

\section{Introduction}
Automatic Music Transcription (AMT) is a field of study which tries to convert acoustic recordings into some form of (digital) music notation~\cite{survey1}. This task can be divided into two subproblems: pitch estimation and onset detection~\cite{survey2}. Instead of trying to convert a song into music notation, this paper will focus on translating a guitar signal into a set of played notes in real-time. This problem is very similar to AMT, as both require pitch estimation and onset/offset detection, however, we also have take latency into account.

Single pitch estimation can be accurately performed using simple mathematical models~\cite{mono}. Multi pitch estimation is much harder as harmonic overtones are difficult to differentiate from the fundamental frequency~\cite{oud}. Because of this, most recent methods use machine learning or artificial intelligence. These methods are not well suited for real-time transcription. Also, no actual (mathematical) understanding of the problem is gained. Because of this, this paper will focus on mathematically solving this problem. Even if the methods researched in this paper might not outperform ML/AI solutions by themselves, they could be used to filter the input data for the ML/AI approaches. This might in turn improve the results of the AI/ML methods significantly.

@@

A big problem in polyphonic pitch estimation is the occurrence of overtones~\cite{oud}. As these overtones also have a periodic nature in the frequency domain, they can be detected using another Fourier transform~\cite{double}. This does however have a big impact on the latency, as performing a Fourier transform is computationally expensive. Also, notes which are at octave intervals have the same periodicity~\cite{octave}, which makes them difficult to distinguish.

Another problem is choosing the real-time constraint for processing the input. Some papers claim 140 ms is sufficient~\cite{sloomboi}. This is however ridiculously long, as it would be about a note difference when playing eight notes in 200 BPM. Also, there already is a large latency due to how audio drivers work~\cite{os}. In practice, musicians often tweak their audio driver settings in order to gain 5 to 10 ms in latency, which makes the difference between being able to play faster pieces or not.

@@

As stated before, our goal is to perform the transcription in real-time. We choose a constraint of 6 ms. Most musicians work with a driver latency of 9 to 11 ms. On a good computer, this latency can be reduced to 3-5 ms. We will use these 6 ms of latency difference to perform our calculations in. Note that when playing more notes per second, latency becomes more of a problem. On slower pieces, we can get away with much larger latencies. In order to test the impact of extra latency for yourself, we developed a tool which plays back the incoming audio signal with a user specifiable delay. This tool can be found under the "latency" branch in our GitHub repository. Note that this tool adds a latency on top of the audio driver latency like our program would. The total latency is much larger than the displayed latency.

@@

\subsection{Transform to frequency domain}
To estimate the frequency components of a frame, transforms are used. Most commonly, the Fourier transform is used as it is the most researched transform and much is known about it~\cite{survey2}. Its main downside is that the frequency bins are constant in size, where the distance between notes increases exponentially. Because of this, low notes are hard to discern in the frequency domain where high notes have more resolution than needed. Because of this, other transforms such as the constant Q transform (CQT) are getting more popular.

When using transforms, window functions have to by applied to the frames to prevent spectral leakage~\cite{window}. Furthermore, the actual peak locations can be accurately interpolated using (Lagrangian) quadratic interpolation~\cite{interpol}. These concepts are further described in Section~\ref{sub:four} and Section~\ref{sub:peak} respectively.

\subsection{Peak picking}
Peaks in the frequency domain correspond to predominant frequencies in the analysed frame. However, not every peak is of interest. Some peaks are generated by noise. Many of these peaks can be filtered by requiring a minimal signal level to be considered a peak. Furthermore, due to spectral leakage, peaks can be observed which are not in the original signal. This kind of noise can be minimized by choose a good window function~\cite{window}. In polyphonic signals, harmonic overtones of different notes may interfere, which will may lead to inconsistent and oscillating peaks.  % TODO: Maybe make preliminary section about overtone dissonance

@@

\section{Implementation}  \label{sec:impl}
The program developed for this paper is written in C++ (11) and compiled with g++ (10.2.0). It uses SDL (2.0.14) for audio input/output and spectrum visualisation. FFTW (3.3.9) is used for efficient Fourier transforms. Its code can be found at "\url{github.com/lucmans/dechord}".

There are many parameters which change the behaviour of the overall system. For this paper, we choose some values which seemed to perform well. The values can still be optimized for better results. All parameters can be found in \texttt{config.h} along with comments explaining them.

@@

Before applying the transform, a window function is often applied to minimize the artifacts produced by the transform. In this paper, we mainly use the Blackman-Nuttall window, but the following windows are also provided: Hamming window, Hann window, Blackman window, Nuttall window, Blackman-Harris window, Flat top window. Other window functions can easily be added.

\subsection{Peak picking} \label{sub:peak}
Three different peak picking algorithms have been implemented. The first picks every local maximum (every \texttt{i} where \texttt{data[i - 1] < data[i] \&\& data[i] > data[i + 1]}). This approach finds too many peaks, especially in noise. The other two algorithms pick peaks based on a Gaussian average envelope. One algorithm picks all peaks which are above this average and the other only picks the highest peak for each spectral lobe above this average. Only picking the highest peak seemed to restrictive in testing, however, it is better resilient against noise and may be used with more careful parameter optimisation. By using the envelope, we effectively filter small local maxima close to large local maxima.

The Gaussian average envelope is calculated for every Fourier bin $i$ by taking the weighted average of a arbitrary number of bins around $i$. The weights are determined by a Gaussian function ($e^{-\pi(\frac{x}{\sigma})^2}$). The value of $\sigma$ and number of bins around $i$ can be changed in \texttt{config.h}. Higher values for $\sigma$ makes points around the considered point have a higher weighting. In our implementation, $\sigma$ is set to 1.2 and the number of bins is set to 47.

Even though this envelope works well for parts of the spectrum with peaks, it will still find peaks in noisy areas without peaks. This can be prevented in two ways. The first is setting a threshold which a peak has to exceed. This will result in less sustain, but works well. Secondly, we know that peaks outside the frequency range of a guitar cannot be a fundamental frequency. We can discard peaks lower than the lowest note, but peaks higher than the highest note still provide us information as it might be an overtone.

@@

We use a frame size of 16384 samples and a sample rate of 192000 samples per second, which equates to a frame length of 85.3 ms. This is well over our real-time constraint. Unfortunately, we cannot change this value too much. The sample rate is already at the maximum of most audio interfaces. We also cannot lower the number of samples per frame while using the Fourier transform, as it would further reduce the frequency resolution. However, as will be explained in Section~\ref{sec:future}, this might not be a problem, as we can reduce this latency by using partially overlapping frames.

@@

We constructed our own dataset to evaluate the performance. This dataset is a technical dataset consisting of scales and intervals played on a guitar. Scales allow us to measure the monophonic performance. As scales are easy to play, multiple recordings should have the same results. This allows us to isolate and measure other variables, such as the effect of tempo or playing the scale higher/lower. Recordings of different intervals allow us to measure the polyphonic performance. By measuring all two note intervals, we can identify any problems with specific intervals. As the distance in frequency in an interval increases when the same interval is played at higher notes, we record the intervals at multiple root notes. We will also measure the performance on the four basic triads: the major triad, minor triad, augmented triad and diminished triad.
